The illusory promise of the Aligned Rank Transform

### 2025-07-04 sandbox.qmd
final corrections on interpretation issues

::: {#fig-effect-main-scatterplots-1}
```{r, echo=FALSE, fig.height=3.4, message=FALSE, warning=FALSE}
source("effect_sizes_scatterplots.R")
prefix <- "6_scatter-Effect_Size-g2"
df <- readlyDataPoints(prefix,  distributions=c("norm", "lnorm", "likert5B"), dnames = c("Normal", "Log-normal", "Ordinal (5 levels)"), effect = "X1X2") %>% toCohensf2()

fig <- plotlyScatter(df, xlab = "Ground truth", ylab= "Cohen's f", xvar = "fX1X2", max = 0.6)

fig
```
Scatterplots showing the relationship between effect size estimates (Cohen's $f$) of each method and estimates of the ground-truth method for the main effect of $X_1$, when $a_1$ and $a_2$ vary within the range $[-2, 2]$. 
:::

>>>>

**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.

By contrast, interpretation is unambiguous in these situations:

- *Main effects*: when the interaction coefficient $a_{12}$ is zero.

- *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.

For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).

Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous. Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately. However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions. Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised. While @elkin:2021 does examine such models, they avoid these scenarios and do not address these problems.

### 2025-07-04 text/40-interpretingEffects.qmd
final corrections on interpretation issues

**Defining interactions on the scale of the response variable.** @Ai:2003 and, more recently, @McCabe:2022 argue that the interaction coefficient alone is not sufficient to describe the actual interaction trend on the natural scale of the response variable. They argue that an interaction effect should instead be defined as the observed change in the marginal effect of $x_1$ as a function of $x_2$. For continuous predictors, this can be formalized as the second-order cross partial derivative of the expected value (which is $\frac{\partial^2 E(Y|x_1, x_2)}{\partial x_1 \partial x_2}$), while for discrete predictors, it can be expressed using discrete differences between levels of the predictors.

Using this definition, the authors show that a zero interaction coefficient ($a_{12} = 0$) does not necessarily imply the absence of an interaction effect. In fact, for continuous predictors, they demonstrate that  when $a_{12} = 0$, the interaction effect is proportional to $a_1$ and $a_2$.

@@

*“Statistical testing about the model specification is done at this step. Hypothesis tests are about model coefficients and about the structural aspects of the model specifications. Partial effects are neither coefficients nor elements of the specification of the model. They are implications of the specified and estimated model”* [@Greene:2010, Page 295].

@@

**Clarifying interpretation boundaries.** Interaction interpretation issues only arise when parallel main effects are present, that is, when the coefficients of all interacting factors are non-zero. But interpretation issues can also affect main effects. In particular, if the interaction coefficient $a_{12}$ is non-zero, we may observe a change in the marginal effect of a factor (e.g., $x_1$) despite the fact that its corresponding model coefficient (e.g., $a_1$) is zero.

We emphasize that the existing literature does not clarify how ART is expected to interpret main and interaction effects in these scenarios. Early evaluations of ART did not consider nonlinear models, so the question was not even posed. @elkin:2021 who do study such models, use log-transformed values as their baseline for log-normal distributions (see their example on Page 4). So it seems that Elkin et al. assume ART's conclusions should align with the coefficients of the linear submodel. Unfortunately, their evaluations focus on contrasts and do not account for parallel effects, avoiding interpretation issues.

Throughout the rest of the article, we make every effort to clarify when the interpretation of an effect is ambiguous. Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately to avoid confusion. Importantly, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* and no interaction. Similarly, in the majority of our experiments demonstrating ART's deficient behavior, we avoid interpretation issues by varying the model coefficient of a single factor while keeping all other coefficients at zero.

>>>>

**Defining interactions on the scale of the response variable.** @Ai:2003 and, more recently, @McCabe:2022 show that the interaction coefficient alone is not sufficient to describe the actual interaction trend on the natural scale of the response variable. They argue that an interaction effect should instead be defined as the observed change in the marginal effect of $x_1$ as a function of $x_2$. Thus, the trends shown in @fig-interactions do represent interactions, since the difference between *Tech* A and *Tech B* changes as a function of *Difficulty* on the original reponse scale (time or perceived performance).

Using this definition, the authors show that a zero interaction coefficient ($a_{12} = 0$) does not necessarily imply the absence of an interaction effect. In fact, for continuous predictors, they demonstrate that when $a_{12} = 0$, the interaction effect can still emerge and is proportional to $a_1$ and $a_2$.

<!--
For continuous predictors, this can be formalized as the second-order cross partial derivative of the expected value (which is $\frac{\partial^2 E(Y|x_1, x_2)}{\partial x_1 \partial x_2}$), while for discrete predictors, it can be expressed using discrete differences between levels of the predictors. -->

@@

> *“Statistical testing about the model specification is done at this step. Hypothesis tests are about model coefficients and about the structural aspects of the model specifications. Partial effects are neither coefficients nor elements of the specification of the model. They are implications of the specified and estimated model”* [@Greene:2010, Page 295].

@@

**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects -- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero.

By contrast, interpretation is unambiguous in these situations:

1. *Main effects*: when the interaction coefficient $a_{12}$ is zero.

2. *Interaction effect*: when either $a_{1}$ or $a_{2}$ is zero.

For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).

Throughout the remainder of the article, we make every effort to clarify when the interpretation of an effect is ambiguous. Because different methods may rely on different assumptions about the null hypothesis, we analyze such situations separately. However, it is important to note that the existing literature offers no guidance on how ART is expected to interpret main and interaction effects under these conditions. Early evaluations of ART did not consider nonlinear models, and therefore this issue was not raised. While @elkin:2021 do examine such models, they do not address interpretation issues.

### 2025-07-04 text/65-Results-EffectSize.qmd
small correction
final corrections on interpretation issues

@fig-eta-squared-2 shows the $\eta^2$ estimates when the effect of $X_1$ is no longer null ($a_1 = 1$). In this scenario, PAR consistently underestimates effect sizes, indicating its unsuitability for such data. RNK also slightly underestimates effects. ART tends to underestimate effects when $a_2 = 0$, but as $a_2$ increases, its estimates become more variable and exhibit an upward bias. INT provides the most accurate and stable estimates, with only a slight deviation when $a_2 = 8$.

@@

We observe that all three rank-based methods struggle to detect large effects, even when the underlying distribution is normal. This result reflects a fundamental limitation of ranks. Consider, for example, two sets of numbers $A= \{1, 2, 3\}$ and $B= \{4, 5, 6\}$. The mean difference between them is $\overline{B} - \overline{A} = 3$, whether calculated from the row values or their ranks. However, if the set $B$ becomes ${9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$. As effects become larger, this problem becomes more apparent.

>>>>

@@

We observe that all three rank-based methods struggle to detect large effects, even when the underlying distribution is normal. This result reflects a fundamental limitation of ranks. Consider, for example, two sets of numbers $A= \{1, 2, 3\}$ and $B= \{4, 5, 6\}$. The mean difference between them is $\overline{B} - \overline{A} = 3$, whether calculated from the row values or their ranks. However, if the set $B$ becomes $\{9, 10, 11\}$, the mean difference of raw values increases to $8$, while the mean difference in ranks remains $3$. As effects become larger, this problem becomes more apparent.
